Rising from the Ashes: A Metallicity-Dependent Star Formation Gap Splits the Milky Way's $\alpha$-Sequences

### 2025-02-13 paper/ms.tex
Move fig 1

\section{Introduction} \label{sec:intro}
\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{figure1.pdf}
  \caption{\textbf{The abundance bimodality seen in the Milky Way can be reproduced in some idealized merger simulations.} In the upper panels, we show the distribution of stars in the \MgFe{}-\FeH{} plane. The lower panels show the distribution of \MgFe{} at a fixed \FeH{} bin of width $0.05\,\dex$. The colors indicate the fixed \FeH values, which are $-0.5$, $-0.25$, $0$, and $0.25$. The left column shows the observed distribution in the Milky Way from ASPCAP DR17 \citep[][J.A.~Holtzman et al., in preparation]{2016AJ....151..144G}, while the right two columns show two idealized merger simulations. The idealized merger simulations are nearly identical, except that in the bimodal simulation the satellite has a starting radius of $129\,\kpc$, while in the unimodal simulation it has a starting radius of $142\,\kpc$. The labels ``unimodal'' and ``bimodal'' are of the \textit{outcome} of the simulation, and do not reflect a particular choice in the setup. The Milky Way (left column) exhibits a strong bimodal distribution of \MgFe{} at various \FeH{}. The idealized merger simulation marked as bimodal (center column) also exhibits a bimodal distribution of \MgFe{}, though the structure is not as strongly defined. The idealized merger simulation marked as unimodal (right column) exhibits only weak structure, if any at all.}
  \label{fig:fig1}
\end{figure*}

% What leads to distribution of heavy elements in a galaxy
Many elements heavier than hydrogen are produced through nuclear fusion in compact objects such as supernovae, dying low mass stars, and neutron star-neutron star mergers \citep[e.g.][]{2023A&ARv..31....1A}. By necessity, stars inherit the constituitive properties of the gas from which they formed. Moreover, the surface abundance of most elements for most stars do not change over most of their lifetime. We therefore have the opportunity to examine the historical record of the gas-phase abundance of a galaxy by way of the distribution of the surface abundances of stars.

@@

In the lower panels of Figure~\ref{fig:fig1}, we show the distributions of \MgFe{} at different fixed \FeH{}. The (blue, orange, red, green) lines show the \MgFe{} distribution at a \FeH{} of ($-0.5$, $-0.25$, $0$, $0.25$), in bins of width $0.05\,\dex$. The distributions of \MgFe{} are offset (but not rescaled) so that they do not overlap. Here, the bimodality seen in the Milky Way is quite striking at lower metallicities. The peaks are well-separated, by $\sim0.2\,\dex$. In the bimodal simulation, the distribution is still clearly bimodal, but the peaks are less well-separated, by $\sim0.1\,\dex$. In the unimodal simulation, there is a hint of some structure at $\FeH \gtrsim 0.25$, but there is not a strong multimodal structure.

>>>>

\section{Introduction} \label{sec:intro}
% What leads to distribution of heavy elements in a galaxy
Many elements heavier than hydrogen are produced through nuclear fusion in compact objects such as supernovae, dying low mass stars, and neutron star-neutron star mergers \citep[e.g.][]{2023A&ARv..31....1A}. By necessity, stars inherit the constituitive properties of the gas from which they formed. Moreover, the surface abundance of most elements for most stars do not change over most of their lifetime. We therefore have the opportunity to examine the historical record of the gas-phase abundance of a galaxy by way of the distribution of the surface abundances of stars.

@@

In the lower panels of Figure~\ref{fig:fig1}, we show the distributions of \MgFe{} at different fixed \FeH{}. The (blue, orange, red, green) lines show the \MgFe{} distribution at a \FeH{} of ($-0.5$, $-0.25$, $0$, $0.25$), in bins of width $0.05\,\dex$. The distributions of \MgFe{} are offset (but not rescaled) so that they do not overlap. Here, the bimodality seen in the Milky Way is quite striking at lower metallicities. The peaks are well-separated, by $\sim0.2\,\dex$. In the bimodal simulation, the distribution is still clearly bimodal, but the peaks are less well-separated, by $\sim0.1\,\dex$. In the unimodal simulation, there is a hint of some structure at $\FeH \gtrsim 0.25$, but there is not a strong multimodal structure.
\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{figure1.pdf}
  \caption{The abundance bimodality seen in the Milky Way can be reproduced in some idealized merger simulations. In the upper panels, we show the distribution of stars in the \MgFe{}-\FeH{} plane. The lower panels show the distribution of \MgFe{} at a fixed \FeH{} bin of width $0.05\,\dex$. The colors indicate the fixed \FeH values, which are $-0.5$, $-0.25$, $0$, and $0.25$. The left column shows the observed distribution in the Milky Way from ASPCAP DR17 \citep[][J.A.~Holtzman et al., in preparation]{2016AJ....151..144G}, while the right two columns show two idealized merger simulations. The idealized merger simulations are nearly identical, except that in the bimodal simulation the satellite has a starting \edit1{velocity of $142\,\kms$}, while in the unimodal simulation it has a starting \edit1{velocity of $116\,\kms$}. The labels ``unimodal'' and ``bimodal'' are of the \textit{outcome} of the simulation, and do not reflect a particular choice in the setup. The Milky Way (left column) exhibits a strong bimodal distribution of \MgFe{} at various \FeH{}. The idealized merger simulation marked as bimodal (center column) also exhibits a bimodal distribution of \MgFe{}, though the structure is not as strongly defined. The idealized merger simulation marked as unimodal (right column) exhibits only weak structure, if any at all.}
  \label{fig:fig1}
\end{figure*}
